---
title: "Data Science Capstone Milestone Report"
author: "Gary Garrison"
date: "March 19, 2015"
output: html_document
---

## Preprocessing
The data for this project was downloaded from https://d396qusza40orc.cloudfront.net/dsscapstone/dataset/Coursera-SwiftKey.zip. The full corpora had the follwing characteristics:

Line, word, and byte counts:
```{r,echo=TRUE}
system("wc  ./final/en_US/* > full_wc.txt")
full_wc <- read.table("full_wc.txt")
names(full_wc) <- c("Lines","Words","Bytes","File.Name")
print(full_wc,row.names = FALSE)
```
The wc command returns the number of lines, words, and bytes for each of the three source files. In the output above the the blogs data has 899,288 lines, 37,334,626 words, and 210,159,817 bytes. This is lot of data so for our model building and exploratory analysis we use only sample of this data

We created a function to sample to get a sample from each of the three types which included news group, blog, and Twitter data. 

After sampling our data had the following characteristics:
```{r, echo=TRUE}
system("wc  ./samples/*")
system("wc  ./samples/* > sample_wc.txt")
sample_wc <- read.table("sample_wc.txt")
names(sample_wc) <- c("Lines","Words","Bytes","File.Name")
print(sample_wc,row.names = FALSE)
```

# Vocabulary size:
udm <- TermDocumentMatrix(crps)

## Exploratory Analysis

### Statistical Summaries


### Exploratory Plots


## Findings

## Plan for Prediction and App Creation


## References

All code for this project can be foundt at the following Github Repoistory:
https://github.com/gwgarrison/data_science_capstone